### 2024-05-20 article/article.tex
henrik placeholder code added

\documentclass[a4paper, twocolumn]{article}

% you can switch between these two (and more) styles by commenting one out (use percentage)
\usepackage[backend=biber]{biblatex}
%\usepackage[backend=biber, style=authoryear-icomp]{biblatex}
\addbibresource{./refs.bib}

\usepackage{graphicx}

\usepackage{listings}
\usepackage{color}
\definecolor{lightgray}{gray}{0.9}

% code listing: https://tex.stackexchange.com/questions/19004/how-to-format-an-inline-source-code
\lstset{
    showstringspaces=false,
    basicstyle=\ttfamily,
    keywordstyle={blue},
    commentstyle=\color[gray]{0.6}
    stringstyle=\color[RGB]{255, 150, 75}
}
\newcommand{\inlinecode}[2]{\colorbox{lightgray}{\lstinline[language=#1]$#2$}}

\author{Henrik Strøm}
\title{Using \LaTeX}

\begin{document}

\twocolumn[
    \begin{@twocolumnfalse}
        \maketitle
        \begin{abstract}
            LaTeX is an extremely powerful typesetting system.
            This document demonstrates some basic usage.
        \end{abstract}
    \end{@twocolumnfalse}
    \vspace{1cm}
]

\section{Introduction\label{sec:Introduction}}
According to \cite{example_website}, ...

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

Here is a reference to a scientific paper, you can find it in the references.
\cite{lecun1995convolutional}
You can do it like this, e.g., if you want to talk about the paper by
\textcite{lecun1995convolutional}
You can also use
\parencite{lecun1995convolutional}

Paragraphs are separated by empty lines.
It might be a good idea to put every sentence on a separate line -- that way \inlinecode{bash}{git} will work on a sentence-by-sentence basis.

\section{Formatting in \LaTeX\label{sec:Formatting in Latex}}

\subsection{Lists}
\label{sec:Lists}

You can make nice lists in \LaTeX:

\begin{itemize}
    \item First item
    \item Second item
\end{itemize}

Or like this:

\begin{enumerate}
    \item First item
        \begin{enumerate}
            \item Sub element
        \end{enumerate}
    \item Second item
        \begin{itemize}
            \item Another sub element
        \end{itemize}
\end{enumerate}

\subsection{Text Decoration\label{sec:Text Decoration}}

Different text decorations like \textit{textit}, \emph{emph} (which might render differently than here), \textbf{textbf} are possible, and it might be a good idea to find a \LaTeX cheat-sheet to find out more.

You can put things in quotes like "Hey, look at that!", or ``Hey, look at that!''.
Notice these are not identical.

\subsection{Quoteblocks\label{sec:Quoteblock}}

Quote blocks look nice:

\begin{quote}
\emph{
I believe that in about fifty years' time it will be possible, to programme computers (..) to make them play the imitation game so well that an average interrogator will not have more than 70 per cent chance of making the right identification after five minutes of questioning.
}
\cite{turing1950computing}
\end{quote}

\subsection{Tables\label{sec:Tables}}

Things like tables are obviously supported in \LaTeX, see table \ref{table:Likert scale and scoring}.

\begin{table}[ht]
    \caption{Likert scale and scoring}
    \centering
    \begin{tabular}{r | l}
        \hline\hline
        Score & Answer \\ [0.5ex] % inserts table %heading
        \hline
        -2 & Strongly disagree \\
        -1 & Disagree to some extent \\
        1 & Agree to some extent \\
        2 & Strongly agree \\ [1ex]
        \hline
    \end{tabular}
    \label{table:Likert scale and scoring}
\end{table}

\subsection{Equations\label{sec:Equations}}

The cost function we most often use in classification is shown in Equation \ref{eq:cost-function-classification}.

\begin{equation}
    \label{eq:cost-function-classification}
    J_\theta(X) = -\frac{1}{m} \sum (Y \cdot log(h_\theta(X)) + (1 - Y) \cdot log(1 - h_\theta(X))
\end{equation}

\subsection{Figures\label{sec:figures}}

Figure \ref{fig:univariate-linear-fit-cost} shows the reduced cost over number of iterations running batch gradient descent.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/univariate-linear-fit-cost.png}
    \caption{Reduction of the cost function over iterations.}
    \label{fig:univariate-linear-fit-cost}
\end{figure}

\printbibliography

\end{document}

### 2024-05-20 article/article.tex
article clean up

@@
 
-\author{Henrik Strøm}
-\title{Using \LaTeX}
+\author{Jeffrey Roed, Rasmus Kibshede, Joachim Richter}
+\title{Genetic algorithms snake game}
 
@@
         \begin{abstract}
-            LaTeX is an extremely powerful typesetting system.
-            This document demonstrates some basic usage.
+            ---------------- TBA --------------------
         \end{abstract}
@@
 \section{Introduction\label{sec:Introduction}}
-According to \cite{example_website}, ...
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+The primary objective of this study is to demonstrate how genetic algorithms (GAs) can be utilized to develop an agent that plays the Snake game.
+The Snake game, a classic arcade game, involves navigating a snake to collect food items while avoiding collisions with walls and the snake's own body.
+The GA is used to optimize the snake's behavior over successive generations, improving its ability to survive and collect food.
 
-Here is a reference to a scientific paper, you can find it in the references.
-\cite{lecun1995convolutional}
-You can do it like this, e.g., if you want to talk about the paper by
-\textcite{lecun1995convolutional}
-You can also use
-\parencite{lecun1995convolutional}
+\subsection{Research question\label{sec:Research Question}}
 
-Paragraphs are separated by empty lines.
-It might be a good idea to put every sentence on a separate line -- that way \inlinecode{bash}{git} will work on a sentence-by-sentence basis.
+How can we develop a genetic algorithm capable of outperforming the average score achieved by our group in the game of Snake?
 
-\section{Formatting in \LaTeX\label{sec:Formatting in Latex}}
+\section{Method\label{sec:Method}}
 
-\subsection{Lists}
-\label{sec:Lists}
-
-You can make nice lists in \LaTeX:
+\subsection{Selection\label{sec:Selection}}
 
 \begin{itemize}
-    \item First item
-    \item Second item
+    \item A selection mechanism is used to choose parent chromosomes for the next generation.
+    \item Techniques such as roulette wheel selection or tournament selection are applied to favor higher fitness individuals.
 \end{itemize}
 
-Or like this:
+\subsection{Crossover\label{sec:Crossover}}
 
-\begin{enumerate}
-    \item First item
-        \begin{enumerate}
-            \item Sub element
-        \end{enumerate}
-    \item Second item
 \begin{itemize}
-            \item Another sub element
+    \item Crossover operators combine pairs of parent chromosomes to produce offspring.
+    \item Single-point or two-point crossover methods are employed to mix the neural network weights and biases of parent agents.
 \end{itemize}
-\end{enumerate}
 
-\subsection{Text Decoration\label{sec:Text Decoration}}
+\subsection{Mutation\label{sec:Mutation}}
 
-Different text decorations like \textit{textit}, \emph{emph} (which might render differently than here), \textbf{textbf} are possible, and it might be a good idea to find a \LaTeX cheat-sheet to find out more.
+\begin{itemize}
+    \item Mutation operators introduce variability by randomly altering parts of an offspring's chromosome.
+    \item A small mutation rate is used to ensure genetic diversity and avoid premature convergence.
+\end{itemize}
 
-You can put things in quotes like "Hey, look at that!", or ``Hey, look at that!''.
-Notice these are not identical.
+\section{Analysis\label{sec:Analysis}}
 
-\subsection{Quoteblocks\label{sec:Quoteblock}}
+\subsection{Selection\label{sec:Selection}}
 
-Quote blocks look nice:
+\subsection{Crossover\label{sec:Crossover}}
 
-\begin{quote}
-\emph{
-I believe that in about fifty years' time it will be possible, to programme computers (..) to make them play the imitation game so well that an average interrogator will not have more than 70 per cent chance of making the right identification after five minutes of questioning.
-}
-\cite{turing1950computing}
-\end{quote}
+\subsection{Mutation\label{sec:Mutation}}
 
-\subsection{Tables\label{sec:Tables}}
+\section{Findings\label{sec:Findings}}
 
-Things like tables are obviously supported in \LaTeX, see table \ref{table:Likert scale and scoring}.
+\subsection{Selection Enhancement\label{sec:Selection Enhancement}}
 
-\begin{table}[ht]
-    \caption{Likert scale and scoring}
-    \centering
-    \begin{tabular}{r | l}
-        \hline\hline
-        Score & Answer \\ [0.5ex] % inserts table %heading
-        \hline
-        -2 & Strongly disagree \\
-        -1 & Disagree to some extent \\
-        1 & Agree to some extent \\
-        2 & Strongly agree \\ [1ex]
-        \hline
-    \end{tabular}
-    \label{table:Likert scale and scoring}
-\end{table}
+\subsection{Crossover Enhancement\label{sec:Crossover Enhancement}}
 
-\subsection{Equations\label{sec:Equations}}
-
-The cost function we most often use in classification is shown in Equation \ref{eq:cost-function-classification}.
-
-\begin{equation}
-    \label{eq:cost-function-classification}
-    J_\theta(X) = -\frac{1}{m} \sum (Y \cdot log(h_\theta(X)) + (1 - Y) \cdot log(1 - h_\theta(X))
-\end{equation}
-
-\subsection{Figures\label{sec:figures}}
-
-Figure \ref{fig:univariate-linear-fit-cost} shows the reduced cost over number of iterations running batch gradient descent.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.5\textwidth]{figures/univariate-linear-fit-cost.png}
-    \caption{Reduction of the cost function over iterations.}
-    \label{fig:univariate-linear-fit-cost}
-\end{figure}
+\subsection{Mutation Enhancement\label{sec:Mutation Enhancement}}
 

### 2024-05-21 article/article.tex
added: new version

@@
 \addbibresource{./refs.bib}
-
 \usepackage{graphicx}
-
 \usepackage{listings}
@@
 \definecolor{lightgray}{gray}{0.9}
-
 % code listing: https://tex.stackexchange.com/questions/19004/how-to-format-an-inline-source-code
@@
 \newcommand{\inlinecode}[2]{\colorbox{lightgray}{\lstinline[language=#1]$#2$}}
-
 \author{Jeffrey Roed, Rasmus Kibshede, Joachim Richter}
@@
 \begin{document}
-
 \twocolumn[
@@
         \begin{abstract}
-            ---------------- TBA --------------------
+            TBA
         \end{abstract}
@@
 
-\section{Introduction\label{sec:Introduction}}
+\section {Introduction\label{sec:Introduction}}
 
@@
 The GA is used to optimize the snake's behavior over successive generations, improving its ability to survive and collect food.
-
 \subsection{Research question\label{sec:Research Question}}
-
 How can we develop a genetic algorithm capable of outperforming the average score achieved by our group in the game of Snake?
@@
 \section{Method\label{sec:Method}}
-
 \subsection{Selection\label{sec:Selection}}
 
+The selection methods is responsible for choosing parent chromosomes that will generate the next generation of individuals.
+Below are some thoughts we have made in choosing the right selection methods for our Snake game.
+
 \begin{itemize}
-    \item A selection mechanism is used to choose parent chromosomes for the next generation.
-    \item Techniques such as roulette wheel selection or tournament selection are applied to favor higher fitness individuals.
+\item \textbf{Roulette Wheel Selection.}
+Individuals are chosen with a probability proportional to their fitness. This maintains genetic diversity and prevents premature convergence \cite{goldberg1991comparative}.
+
+Using this method early in the training process is advantageous when the fitness landscape is smooth and diverse strategies need to be explored.
+
+\item \textbf{Tournament Selection.}
+Randomly selected individuals compete, and the winner is chosen for breeding.
+
+Will be useful when the algorithm progresses and a more refined selection is needed to accelerate the optimization process \cite{goldberg1991comparative}.
+
+\item \textbf{Rank Selection.}
+Individuals are ranked based on fitness values, and selection probabilities are based on ranks. This reduces the risk of premature convergence.
+
+Will be useful throughout the evolutionary process by providing steady selection pressure and maintaining diversity \cite{srinvas1994adaptive}.
+
 \end{itemize}
 
+To determine the most effective selection method for our GA, we propose to evaluate each approach by conducting multiple runs of the Snake game simulation.
+The performance will be assessed based on the average score achieved by the evolved agents over a fixed number of generations.
+This will help us determine the most suitable strategy for our Snake game agent.
+Adaptive approaches to selection, as discussed by Srinivas and Patnaik (1994), may also be considered to enhance the overall performance and robustness of the algorithm \cite{srinvas1994adaptive}.
+
+\subsection{Fitness\label{sec:Fitness}}
+
 \subsection{Crossover\label{sec:Crossover}}
+The crossover methods are used in GAs to combine the genetic information of two parent chromosomes in order to generate new offspring.
+Below are some thoughts we have made in choosing the right crossover methods for our Snake game.
 
 \begin{itemize}
-    \item Crossover operators combine pairs of parent chromosomes to produce offspring.
-    \item Single-point or two-point crossover methods are employed to mix the neural network weights and biases of parent agents.
+    \item \textbf{Single point Crossover}
+A random crossover point is selected on the parent chromosomes.
+All of the genetic information is copied from one parent, and all the genetic information after this point is copied from the other parent.
+This is a  simple method that keeps the genetic information of both parents making is useful for upholding diversity in the population.
+
+This will be beneficial in the early stages of training, when we need to explore diverse strategies. 
 \end{itemize}
+
+\begin{itemize}
+    \item \textbf{Two-Point Crossover}
+Selects two random crossover point on the parent chromosomes.
+The genetic information is interchanged between the parents to create two new offspring.
+This method provides a higher level of mixing compared to single-point crossover, which can lead to more varied breeding.
+
+This will be useful in later stages of training, where we need to fine tune the algorithm strategies by combining the most desirable features of parents. 
+\end{itemize}
+
+As highlighted by Henrik Strøm \cite{strom2024travelling}, the biggest challenge in implementing crossover for problems such as Traveling Salesman Problem (TSP), is building an effective crossover mechanism.
+TSP is about finding the shortest possible route to visit a number of cities only once,and afterwards return to the starting city.
+Similar considerations apply to our Snake game GA, where the crossover method must effectively combine parent strategies without breaking the underlying structure of the Nerual Network (NN).
+
+In order to address these challenges, we will also explore crossover methods designed for TSP, such as; Order Crossover, and Partially Mapped Crossover, which maintains validity of the offspring.
+We will during our research experiment with different techniques in order to find the most optimal crossover method.
 
@@
 
-\begin{itemize}
-    \item Mutation operators introduce variability by randomly altering parts of an offspring's chromosome.
-    \item A small mutation rate is used to ensure genetic diversity and avoid premature convergence.
-\end{itemize}
+The mutation function aims to modify the genome of the offspring.
+These modifications serve to maintain and introduce diversity in the genetic population, which is crucial for the overall success of the genetic algorithm.
+\begin{quote}
+\emph{
+    The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions.
+    }
+    \cite{srinvas1994adaptive}
+\end{quote}
 
 \section{Analysis\label{sec:Analysis}}
-
 \subsection{Selection\label{sec:Selection}}
-
 \subsection{Crossover\label{sec:Crossover}}
-
 \subsection{Mutation\label{sec:Mutation}}
@@
 \section{Findings\label{sec:Findings}}
-
 \subsection{Selection Enhancement\label{sec:Selection Enhancement}}
-
 \subsection{Crossover Enhancement\label{sec:Crossover Enhancement}}
-
 \subsection{Mutation Enhancement\label{sec:Mutation Enhancement}}
-
 \printbibliography
-
 \end{document}

### 2024-05-25 article/article.tex
added the new version

@@
 \subsection{Fitness\label{sec:Fitness}}
+The fitness function evaluates how well each individual in a population solves the given problem, assigning a numerical score based on their performance.
+This score guides the selection process for reproduction, favoring higher-scoring individuals to pass their traits to the next generation.
+By driving the evolutionary progress and maintaining diversity, the fitness function helps the algorithm converge toward optimal or near-optimal solutions
 
@@
 The mutation function aims to modify the genome of the offspring.
-These modifications serve to maintain and introduce diversity in the genetic population, which is crucial for the overall success of the genetic algorithm.
+These modifications serve to maintain and introduce diversity in the genetic population, which is crucial for the overall success of the genetic algorithm, and prevents premature convergence.
 \begin{quote}
 \emph{
-    The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions.
+    "The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions."
     }

### 2024-05-26 ga_controller.py
complete GA Snake game

@@ -1,0 +1,82 @@
+from vector import Vector
+import pygame
+from game_controller import GameController
+
+class GAController(GameController):
+    def __init__(self, game, model, display=False):
+        self.display = display
+        self.game = game
+        self.model = model
+        self.action_space = (Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0))
+        self.current_direction = self.action_space[0]  # Initialize with any direction (e.g., up)
+        self.screen = None
+        if self.display:
+            pygame.init()
+            self.screen = pygame.display.set_mode((game.grid.x * game.scale, game.grid.y * game.scale))
+            self.clock = pygame.time.Clock()
+            self.color_snake_head = (0, 255, 0)
+            self.color_food = (255, 0, 0)
+
+    def __del__(self):
+        if self.display and pygame.get_init():
+            pygame.quit()
+
+    def update(self) -> Vector:
+        # observation space
+        dn = self.game.snake.p.y / self.game.grid.y  # Normalize by grid height
+        de = (self.game.grid.x - self.game.snake.p.x) / self.game.grid.x  # Normalize by grid width
+        ds = (self.game.grid.y - self.game.snake.p.y) / self.game.grid.y  # Normalize by grid height
+        dw = self.game.snake.p.x / self.game.grid.x  # Normalize by grid width
+
+        # Normalized delta food x and y
+        dfx = (self.game.food.p.x - self.game.snake.p.x) / self.game.grid.x
+        dfy = (self.game.food.p.y - self.game.snake.p.y) / self.game.grid.y
+
+        # Additional observations for nearby obstacles (e.g., snake's body segments)
+        left_obstacle = int(self.game.snake.p.x == 0)
+        right_obstacle = int(self.game.snake.p.x == self.game.grid.x - 1)
+        up_obstacle = int(self.game.snake.p.y == 0)
+        down_obstacle = int(self.game.snake.p.y == self.game.grid.y - 1)
+
+        # Include observations about the presence of obstacles in each direction
+        obs = (dn, de, ds, dw, dfx, dfy, left_obstacle, right_obstacle, up_obstacle, down_obstacle)
+
+        if len(self.game.snake.body) >= 2:
+            tail_left = int(self.game.snake.p - self.game.snake.body[1] == Vector(-1, 0))
+            tail_right = int(self.game.snake.p - self.game.snake.body[1] == Vector(1, 0))
[…]

### 2024-05-26 ga_models/activation.py

@@ -1,0 +1,22 @@
+import numpy as np
+
+
+def softmax(z):
+    return np.exp(z) / np.sum(np.exp(z))
+
+
+def sigmoid(z):
+    return 1 / (1 + np.exp(-z))
+
+
+def tanh(z):
+    return np.tanh(z)
+
+
+def relu(z):
+    return np.maximum(0, z)
+
+
+def leaky_relu(z):
+    epsilon = 0.001
+    return np.maximum(z * epsilon, z)

### 2024-05-26 ga_models/ga_simple.py

@@ -1,0 +1,64 @@
+import random
+from typing import Protocol, Tuple, List, Sequence
+import numpy as np
+from ga_models.ga_protocol import GAModel
+from ga_models.activation import sigmoid, tanh, softmax
+
+
+class SimpleModel(GAModel):
+    def __init__(self, *, dims: Tuple[int, ...]):
+        assert len(dims) >= 2, 'Error: dims must be two or higher.'
+        self.dims = dims
+        self.DNA = []
+        self.activation_functions = [sigmoid for _ in range(len(dims) - 2)] + [tanh]
+        self.DNA = [np.random.rand(dim, dims[i+1]) for i, dim in enumerate(dims[:-1])]
+
+    def update(self, obs: Sequence) -> Tuple[int, ...]:
+        x = obs
+        for layer, activation in zip(self.DNA, self.activation_functions):
+            x = activation(np.dot(x, layer))
+        return softmax(x)
+
+    def action(self, obs: Sequence):
+        # Prioritize actions that move towards food and away from obstacles
+        if obs[4] < 0:  # Food is to the left
+            return 3  # Move left
+        elif obs[4] > 0:  # Food is to the right
+            return 2  # Move right
+        elif obs[5] < 0:  # Food is above
+            return 0  # Move up
+        elif obs[5] > 0:  # Food is below
+            return 1  # Move down
+        elif obs[8]:  # Tail is to the left
+            return 3  # Move left
+        elif obs[9]:  # Tail is to the right
+            return 2  # Move right
+        elif obs[10]:  # Tail is above
+            return 0  # Move up
+        elif obs[11]:  # Tail is below
+            return 1  # Move down
+        else:
+            # If no food or tail is nearby, choose a random action
+            return random.randint(0, 3)
+
+    def mutate(self, mutation_rate) -> None:
+        if random.random() < mutation_rate:
+            random_layer = random.randint(0, len(self.DNA) - 1)
+            row = random.randint(0, self.DNA[random_layer].shape[0] - 1)
+            col = random.randint(0, self.DNA[random_layer].shape[1] - 1)
+            self.DNA[random_layer][row][col] = random.uniform(-1, 1)
+
+    def __add__(self, other):
+        baby_DNA = []
+        for mom, dad in zip(self.DNA, other.DNA):
+            if random.random() > 0.5:
+                baby_DNA.append(mom)
+            else:
+                baby_DNA.append(dad)
[…]

### 2024-05-26 ga_snake.py

@@ -1,0 +1,89 @@
+import random
+from ga_models.ga_simple import SimpleModel
+from ga_controller import GAController
+from snake import SnakeGame
+
+# Define the dimensions for the neural network
+input_size = 11  # Number of features in the observation space
+hidden_layer_size = 16  # Number of neurons in the hidden layer
+output_size = 4  # Number of possible actions
+
+# Parameters for the GA
+population_size = 50
+mutation_rate = 0.05
+num_generations = 100
+
+# Initialize the population
+population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size)) for _ in range(population_size)]
+
+
+# Define the fitness function
+def evaluate_fitness(model):
+    game = SnakeGame(controller=None)  # Initialize the game without a controller first
+    controller = GAController(game=game, model=model, display=False)
+    game.controller = controller  # Assign the controller after initialization
+    game.run()  # Run the game
+    return game.snake.score
+
+
+# Run the GA
+for generation in range(num_generations):
+    fitness_scores = []
+    for idx, individual in enumerate(population):
+        score = evaluate_fitness(individual)
+        fitness_scores.append((score, individual))
+
+    # Sort individuals based on fitness scores
+    fitness_scores.sort(key=lambda x: x[0], reverse=False)
+
+    # Select the best individuals
+    selected_individuals = [individual for _, individual in fitness_scores[:population_size // 2]]
+
+    # Create the next generation
+    next_generation = []
+    for _ in range(population_size):
+        parent1, parent2 = random.sample(selected_individuals, 2)
+        offspring = parent1 + parent2
+        offspring.mutate(mutation_rate)
+        next_generation.append(offspring)
+
+    # Replace the old population with the new one
+    population = next_generation
+
+    # Print the average fitness of the generation
+    average_fitness = sum(score for score, _ in fitness_scores) / population_size
+    print(f'Generation {generation + 1}: Average Fitness = {average_fitness}')
+
+# Run the GA
+for generation in range(num_generations):
+    fitness_scores = []
+    for idx, individual in enumerate(population):
+        score = evaluate_fitness(individual)
+        fitness_scores.append(score)
+        #print(f'Individual {idx + 1}: Fitness = {score}')
+
+    # Print scores for debugging
+    #print(f'Generation {generation + 1} scores: {fitness_scores}')
+
+    # Select the best individuals
+    selected_individuals = random.choices(
+        population, weights=fitness_scores, k=population_size // 2)
+
[…]

### 2024-05-26 game_controller.py

@@ -1,0 +1,52 @@
+from typing import Protocol
+from vector import Vector
+import pygame
+
+
+class GameController(Protocol):
+    def update(self) -> Vector:
+        print('fake Controller')
+        pass
+
+
+class HumanController(GameController):
+    def __init__(self, game):
+        self.game = game
+        self.game.controller = self
+        pygame.init()
+        self.screen = pygame.display.set_mode((game.grid.x * game.scale, game.grid.y * game.scale))
+        self.clock = pygame.time.Clock()
+        self.color_snake_head = (0, 255, 0)
+        self.color_food = (255, 0, 0)
+
+    def __del__(self):
+        pygame.quit()
+
+    def update(self) -> Vector:
+        next_move = None
+        for event in pygame.event.get():
+            if event.type == pygame.QUIT:
+                running = False
+            if event.type == pygame.KEYDOWN:
+                if event.key == pygame.K_LEFT:
+                    next_move = Vector(-1, 0)
+                if event.key == pygame.K_RIGHT:
+                    next_move = Vector(1, 0)
+                if event.key == pygame.K_UP:
+                    next_move = Vector(0, -1)
+                if event.key == pygame.K_DOWN:
+                    next_move = Vector(0, 1)
+        self.screen.fill('black')
+        for i, p in enumerate(self.game.snake.body):
+            pygame.draw.rect(self.screen, (0, max(128, 255 - i * 12), 0), self.block(p))
+        pygame.draw.rect(self.screen, self.color_food, self.block(self.game.food.p))
+        pygame.display.flip()
+        self.clock.tick(10)
+        print(next_move)
+        return next_move
+
+    def block(self, obj):
+        return (obj.x * self.game.scale,
+                obj.y * self.game.scale,
+                self.game.scale,
+                self.game.scale)

### 2024-05-26 snake.py

@@ -1,0 +1,73 @@
+from collections import deque
+from ga_controller import GAController
+from vector import Vector
+
+class SnakeGame:
+    def __init__(self, xsize: int=30, ysize: int=30, scale: int=15, controller=None):
+        self.grid = Vector(xsize, ysize)
+        self.scale = scale
+        self.snake = Snake(game=self)
+        self.food = Food(game=self)
+        self.controller = controller  # Add this line to accept a controller
+
+    def run(self):
+        running = True
+        while running:
+            next_move = self.controller.update()
+            if next_move:
+                self.snake.v = next_move
+                self.snake.move()
+            if not self.snake.p.within(self.grid):
+                running = False
+                #message = 'Game over! You crashed into the wall!'
+            if self.snake.cross_own_tail:
+                running = False
+                #message = 'Game over! You hit your own tail!'
+            if self.snake.p == self.food.p:
+                self.snake.add_score()
+                self.food = Food(game=self)
+        #print(f'{message} ... Score: {self.snake.score}')
+
+class Food:
+    def __init__(self, game: SnakeGame):
+        self.game = game
+        self.p = Vector.random_within(self.game.grid)
+
+class Snake:
+    def __init__(self, *, game: SnakeGame):
+        self.game = game
+        self.score = 0
+        self.v = Vector(0, 0)
+        self.body = deque()
+        self.body.append(Vector.random_within(self.game.grid))
+
+    def move(self):
+        self.p = self.p + self.v
+
+    @property
+    def cross_own_tail(self):
+        try:
+            self.body.index(self.p, 1)
+            return True
+        except ValueError:
+            return False
+
+    @property
+    def p(self):
+        return self.body[0]
+
+    @p.setter
+    def p(self, value):
+        self.body.appendleft(value)
+        self.body.pop()
+
+    def add_score(self):
+        self.score += 1
+        tail = self.body.pop()
+        self.body.append(tail)
+        self.body.append(tail)
+
+    def debug(self):
+        print('===')
+        for i in self.body:
+            print(str(i))

### 2024-05-26 snake_game.py

@@ -1,0 +1,9 @@
+
+from snake import SnakeGame
+from game_controller import HumanController
+
+
+if __name__ == '__main__':
+    game = SnakeGame()
+    controller = HumanController(game)
+    game.run()

### 2024-05-26 vector.py

@@ -1,0 +1,28 @@
+import random
+
+class Vector:
+    def __init__(self, x: int = 0, y: int = 0):
+        self.x = x
+        self.y = y
+
+    def __str__(self):
+        return f'Vector({self.x}, {self.y})'
+
+    def __add__(self, other: 'Vector') -> 'Vector':
+        return Vector(self.x + other.x, self.y + other.y)
+
+    def within(self, scope: 'Vector') -> bool:
+        return self.x <= scope.x and self.x >= 0 and self.y <= scope.y and self.y >= 0
+
+    def __eq__(self, other: 'Vector') -> bool:
+        return self.x == other.x and self.y == other.y
+
+    def __neg__(self):
+        return Vector(-self.x, -self.y)
+
+    def __sub__(self, other: 'Vector') -> 'Vector':
+        return Vector(self.x - other.x, self.y - other.y)
+
+    @classmethod
+    def random_within(cls, scope: 'Vector') -> 'Vector':
+        return Vector(random.randint(0, scope.x - 1), random.randint(0, scope.y - 1))

### 2024-05-28 article/article.tex
added the new version
added mutation v2

@@
 \subsection{Research question\label{sec:Research Question}}
-How can we develop a genetic algorithm capable of outperforming the average score achieved by our group in the game of Snake?
+How can we develop a genetic algorithm that not only matches but exceeds our group's average score in the game of Snake by leveraging evolutionary strategies, including effective initialization, precise fitness evaluation, robust selection, and strategic use of crossover and mutation?
 
@@
 
+\subsection{Mutation V.2 Jeff\label{sec:Mutation}}
+
+Mutation is a critical component of genetic algorithms (GAs), serving to introduce genetic diversity into the population and prevent premature convergence to suboptimal solutions.
+By altering the genetic material of offspring, mutation helps maintain a healthy level of variability, ensuring the exploration of a broader solution space.
+
+David E.
+Goldberg, in his foundational text \textit{Genetic Algorithms in Search, Optimization, and Machine Learning}, emphasizes the importance of mutation in maintaining genetic diversity and aiding the convergence of the algorithm to optimal solutions.
+He notes that mutation plays a crucial role in restoring lost or unexplored genetic material, thus enhancing the robustness of GAs \cite{goldberg1989genetic}.
+
+Robert E.
+Marks, in \textit{Playing Games with Genetic Algorithms}, highlights the application of GAs in game strategies, noting that mutation is essential for evolving strategies that can adapt to dynamic environments and unforeseen challenges \cite{marks2001playing}.
+
+\textbf{Types of Mutation:}
+\begin{itemize}
+    \item \textbf{Bit Flip Mutation:} Common in binary-encoded GAs, this method involves flipping one or more bits in a chromosome.
+Simple yet effective for maintaining genetic diversity.
+Suitable for early stages of evolution where broad exploration is needed.
+    \item \textbf{Swap Mutation:} Involves swapping two genes within a chromosome. Particularly useful for problems like the Traveling Salesman Problem (TSP) and similar permutation-based problems.
+    \item \textbf{Scramble Mutation:} Selects a subset of genes and randomly scrambles them. Maintains the overall genetic material but alters the order, fostering diversity.
+    \item \textbf{Gaussian Mutation:} Applied to real-valued GAs, where a small Gaussian noise is added to the genes. Effective for fine-tuning solutions in continuous optimization problems.
+\end{itemize}
+
+\begin{quote}
+\emph{"The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions."} \cite{srinvas1994adaptive}
+\end{quote}
+
+\textbf{Implementing Mutation in the Snake Game GA:} To enhance the performance of our Snake game agent, we propose to experiment with various mutation techniques, monitoring their impact on the agent's ability to adapt and improve over successive generations.
+By fine-tuning the mutation rate and method, we aim to balance exploration and exploitation, leading to a more robust and high-performing Snake game strategy.
+
 \section{Analysis\label{sec:Analysis}}
 \subsection{Selection\label{sec:Selection}}
+\subsection{Fitness\label{sec:Fitness}}
 \subsection{Crossover\label{sec:Crossover}}

### 2024-05-29 ga_controller.py
new changes

@@ -74,7 +74,7 @@ class GAController(GameController):
                     pygame.draw.rect(self.screen, (0, max(128, 255 - i * 12), 0), self.block(p))
                 pygame.draw.rect(self.screen, self.color_food, self.block(self.game.food.p))
                 pygame.display.flip()
-                self.clock.tick(10)
+                self.clock.tick(40)
 
         return next_move
 

### 2024-05-29 article/article.tex
new changes

@@
 
-The mutation function aims to modify the genome of the offspring.
-These modifications serve to maintain and introduce diversity in the genetic population, which is crucial for the overall success of the genetic algorithm, and prevents premature convergence.
-\begin{quote}
-\emph{
-    "The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions."
-    }
-    \cite{srinvas1994adaptive}
-\end{quote}
-
-\subsection{Mutation V.2 Jeff\label{sec:Mutation}}
-
 Mutation is a critical component of genetic algorithms (GAs), serving to introduce genetic diversity into the population and prevent premature convergence to suboptimal solutions.
@@
 \subsection{Mutation\label{sec:Mutation}}
+single point mutation
 

### 2024-05-29 ga_controller.py
neural network implemented instead of hardcoded actions

@@ -2,8 +2,9 @@ from vector import Vector
 import pygame
 from game_controller import GameController
 
+
 class GAController(GameController):
-    def __init__(self, game, model, display=False):
+    def __init__(self, game, model, display=True):
         self.display = display
         self.game = game
         self.model = model

### 2024-05-29 ga_models/ga_simple.py

@@ -10,53 +10,37 @@ class SimpleModel(GAModel):
         assert len(dims) >= 2, 'Error: dims must be two or higher.'
         self.dims = dims
         self.DNA = []
-        self.activation_functions = [sigmoid for _ in range(len(dims) - 2)] + [tanh]
+        self.activation_functions = [tanh for _ in range(len(dims) - 2)] + [sigmoid]
         self.DNA = [np.random.rand(dim, dims[i+1]) for i, dim in enumerate(dims[:-1])]
 
     def update(self, obs: Sequence) -> Tuple[int, ...]:
         x = obs
         for layer, activation in zip(self.DNA, self.activation_functions):
             x = activation(np.dot(x, layer))
-        return softmax(x)
+        return sigmoid(x)
 
     def action(self, obs: Sequence):
-        # Prioritize actions that move towards food and away from obstacles
-        if obs[4] < 0:  # Food is to the left
-            return 3  # Move left
-        elif obs[4] > 0:  # Food is to the right
-            return 2  # Move right
-        elif obs[5] < 0:  # Food is above
-            return 0  # Move up
-        elif obs[5] > 0:  # Food is below
-            return 1  # Move down
-        elif obs[8]:  # Tail is to the left
-            return 3  # Move left
-        elif obs[9]:  # Tail is to the right
-            return 2  # Move right
-        elif obs[10]:  # Tail is above
-            return 0  # Move up
-        elif obs[11]:  # Tail is below
-            return 1  # Move down
-        else:
-            # If no food or tail is nearby, choose a random action
-            return random.randint(0, 3)
+        # Use the neural network to predict the action
+        output = self.update(obs)
+
+        # Choose the action corresponding to the highest output neuron
+        return np.argmax(output)
 
     def mutate(self, mutation_rate) -> None:
-        if random.random() < mutation_rate:
-            random_layer = random.randint(0, len(self.DNA) - 1)
-            row = random.randint(0, self.DNA[random_layer].shape[0] - 1)
-            col = random.randint(0, self.DNA[random_layer].shape[1] - 1)
-            self.DNA[random_layer][row][col] = random.uniform(-1, 1)
+        for layer_idx, layer in enumerate(self.DNA):
+            if random.random() < mutation_rate:
+                for row_idx in range(layer.shape[0]):
+                    for col_idx in range(layer.shape[1]):
[…]

### 2024-05-29 ga_snake.py

@@ -1,17 +1,20 @@
 import random
+
+import pygame
+
 from ga_models.ga_simple import SimpleModel
 from ga_controller import GAController
 from snake import SnakeGame
 
 # Define the dimensions for the neural network
-input_size = 11  # Number of features in the observation space
-hidden_layer_size = 16  # Number of neurons in the hidden layer
+input_size = 14  # Number of features in the observation space
+hidden_layer_size = 64  # Number of neurons in the hidden layer
 output_size = 4  # Number of possible actions
 
 # Parameters for the GA
-population_size = 50
-mutation_rate = 0.05
-num_generations = 100
+population_size = 5
+mutation_rate = 0.9  # Adjust mutation rate
+num_generations = 10
 
 # Initialize the population
 population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size)) for _ in range(population_size)]
@@ -23,11 +26,49 @@ def evaluate_fitness(model):
     controller = GAController(game=game, model=model, display=False)
     game.controller = controller  # Assign the controller after initialization
     game.run()  # Run the game
-    return game.snake.score
+
+    # Reward for eating food
+    food_reward = 100 * game.snake.score
+
+    # Penalty for hitting walls
+    wall_collision_penalty = -10 if not game.snake.p.within(game.grid) else 0
+
+    # Distance to food
+    distance_to_food = 1 / (1 + game.snake.distance_to_food())
+
+    # Reward for moving towards food
+    food_direction_reward = 10 if game.snake.p + game.snake.v == game.food.p else 0
+
+    # Reward for avoiding wall collision
+    avoid_wall_reward = 5 if game.snake.p.within(game.grid) else 0
+
+    # Reward for changing direction (not going in a straight line)
+    change_direction_reward = 2 if game.snake.v != controller.current_direction else 0
+
+    # Calculate fitness based on score, distance to food, and rewards/penalties
+    fitness = (game.snake.score + food_reward +
+               wall_collision_penalty + distance_to_food +
+               food_direction_reward + avoid_wall_reward + change_direction_reward)
+
+    # Reward for approaching food
+    approach_food_reward = 10 if game.snake.distance_to_food() < 1 else 0  # Reward for getting closer to food
+
+    # Reward for exploring new areas
+    exploration_reward = 0.1 if game.snake.p not in game.snake.body else 0  # Reward for moving to unvisited areas
+
+    # Update fitness with additional rewards
+    fitness += approach_food_reward + exploration_reward
+
[…]

### 2024-05-29 snake.py

@@ -1,12 +1,18 @@
 from collections import deque
 from ga_controller import GAController
 from vector import Vector
+import random
+from math import sqrt
+
 
 class SnakeGame:
-    def __init__(self, xsize: int=30, ysize: int=30, scale: int=15, controller=None):
+    def __init__(self, xsize: int=30, ysize: int=30, scale: int=15, max_steps: int=1000, controller=None):
         self.grid = Vector(xsize, ysize)
         self.scale = scale
-        self.snake = Snake(game=self)
+        self.max_steps = max_steps  # Maximum allowed steps
+        self.current_step = 0  # Initialize step counter
+        initial_direction = random.choice([Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)])
+        self.snake = Snake(game=self, initial_direction=initial_direction)
         self.food = Food(game=self)
         self.controller = controller  # Add this line to accept a controller
 
@@ -17,12 +23,13 @@ class SnakeGame:
             if next_move:
                 self.snake.v = next_move
                 self.snake.move()
+                self.current_step += 1  # Increment step counter
+                if self.current_step >= self.max_steps:
+                    running = False  # Terminate the game if max step count reached
             if not self.snake.p.within(self.grid):
                 running = False
-                #message = 'Game over! You crashed into the wall!'
             if self.snake.cross_own_tail:
                 running = False
-                #message = 'Game over! You hit your own tail!'
             if self.snake.p == self.food.p:
                 self.snake.add_score()
                 self.food = Food(game=self)
@@ -34,15 +41,23 @@ class Food:
         self.p = Vector.random_within(self.game.grid)
 
 class Snake:
-    def __init__(self, *, game: SnakeGame):
+    def __init__(self, *, game: SnakeGame, initial_direction: Vector = None):
         self.game = game
         self.score = 0
-        self.v = Vector(0, 0)
+        if initial_direction:
+            self.v = initial_direction
+        else:
+            self.v = random.choice([Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)])  # Randomize initial direction
         self.body = deque()
         self.body.append(Vector.random_within(self.game.grid))
 
+
     def move(self):
         self.p = self.p + self.v
+
+    def distance_to_food(self):
+        # Calculate Euclidean distance between snake's head and food
+        return sqrt((self.game.food.p.x - self.p.x) ** 2 + (self.game.food.p.y - self.p.y) ** 2)
 
     @property
     def cross_own_tail(self):

### 2024-05-29 ga_models/ga_simple.py
added save model and load model

@@ -3,6 +3,7 @@ from typing import Protocol, Tuple, List, Sequence
 import numpy as np
 from ga_models.ga_protocol import GAModel
 from ga_models.activation import sigmoid, tanh, softmax
+import pickle
 
 
 class SimpleModel(GAModel):
@@ -45,4 +46,13 @@ class SimpleModel(GAModel):
 
     def DNA(self):
         return self.DNA
+
+    def save(self, filename):
+        with open(filename, 'wb') as file:
+            pickle.dump(self, file)
+
+    @staticmethod
+    def load(filename):
+        with open(filename, 'rb') as file:
+            return pickle.load(file)
 

### 2024-05-29 ga_snake.py

@@ -1,7 +1,5 @@
 import random
-
 import pygame
-
 from ga_models.ga_simple import SimpleModel
 from ga_controller import GAController
 from snake import SnakeGame
@@ -19,7 +17,6 @@ num_generations = 10
 # Initialize the population
 population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size)) for _ in range(population_size)]
 
-
 # Define the fitness function
 def evaluate_fitness(model):
     game = SnakeGame(controller=None)  # Initialize the game without a controller first
@@ -34,30 +31,22 @@ def evaluate_fitness(model):
     wall_collision_penalty = -10 if not game.snake.p.within(game.grid) else 0
 
     # Distance to food
-    distance_to_food = 1 / (1 + game.snake.distance_to_food())
-
-    # Reward for moving towards food
-    food_direction_reward = 10 if game.snake.p + game.snake.v == game.food.p else 0
-
-    # Reward for avoiding wall collision
-    avoid_wall_reward = 5 if game.snake.p.within(game.grid) else 0
-
-    # Reward for changing direction (not going in a straight line)
-    change_direction_reward = 2 if game.snake.v != controller.current_direction else 0
-
-    # Calculate fitness based on score, distance to food, and rewards/penalties
-    fitness = (game.snake.score + food_reward +
-               wall_collision_penalty + distance_to_food +
-               food_direction_reward + avoid_wall_reward + change_direction_reward)
+    distance_to_food_reward = 1 / (1 + game.snake.distance_to_food())
 
     # Reward for approaching food
-    approach_food_reward = 10 if game.snake.distance_to_food() < 1 else 0  # Reward for getting closer to food
+    approach_food_reward = 10 if game.snake.distance_to_food() < 1 else 0
 
     # Reward for exploring new areas
-    exploration_reward = 0.1 if game.snake.p not in game.snake.body else 0  # Reward for moving to unvisited areas
+    exploration_reward = 0.1 if game.snake.p not in game.snake.body else 0
 
-    # Update fitness with additional rewards
-    fitness += approach_food_reward + exploration_reward
+    # Calculate fitness based on score, distance to food, and rewards/penalties
+    fitness = (
+        food_reward +
+        wall_collision_penalty +
+        distance_to_food_reward +
+        approach_food_reward +
+        exploration_reward
+    )
 
     # Add a baseline value to ensure positive fitness scores
     baseline = max(0, -wall_collision_penalty)
[…]

### 2024-05-30 ga_controller.py
finetuning model

@@ -1,6 +1,8 @@
+import random
 from vector import Vector
 import pygame
 from game_controller import GameController
+import numpy as np
 
 
 class GAController(GameController):
@@ -11,6 +13,7 @@ class GAController(GameController):
         self.action_space = (Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0))
         self.current_direction = self.action_space[0]  # Initialize with any direction (e.g., up)
         self.screen = None
+        self.direction_history = []
         if self.display:
             pygame.init()
             self.screen = pygame.display.set_mode((game.grid.x * game.scale, game.grid.y * game.scale))
@@ -29,7 +32,7 @@ class GAController(GameController):
         ds = (self.game.grid.y - self.game.snake.p.y) / self.game.grid.y  # Normalize by grid height
         dw = self.game.snake.p.x / self.game.grid.x  # Normalize by grid width
 
-        # Normalized delta food x and y
+        # Normalized delta food x and y (food's relative coordinates)
         dfx = (self.game.food.p.x - self.game.snake.p.x) / self.game.grid.x
         dfy = (self.game.food.p.y - self.game.snake.p.y) / self.game.grid.y
 
@@ -40,7 +43,7 @@ class GAController(GameController):
         down_obstacle = int(self.game.snake.p.y == self.game.grid.y - 1)
 
         # Include observations about the presence of obstacles in each direction
-        obs = (dn, de, ds, dw, dfx, dfy, left_obstacle, right_obstacle, up_obstacle, down_obstacle)
+        obs = [dn, de, ds, dw, dfx, dfy, left_obstacle, right_obstacle, up_obstacle, down_obstacle]
 
         if len(self.game.snake.body) >= 2:
             tail_left = int(self.game.snake.p - self.game.snake.body[1] == Vector(-1, 0))
@@ -51,16 +54,48 @@ class GAController(GameController):
             # Default values if snake's body has fewer than two segments
             tail_left, tail_right, tail_up, tail_down = 0, 0, 0, 0
 
-            # Include observations about the presence of the tail in each direction
-        obs += (tail_left, tail_right, tail_up, tail_down)
+        # Include observations about the presence of the tail in each direction
+        obs += [tail_left, tail_right, tail_up, tail_down]
 
-        next_move = self.action_space[self.model.action(obs)]
+        # Possible routes to food
+        route_x_first = [Vector(dfx, 0), Vector(0, dfy)]
[…]

### 2024-05-30 ga_models/ga_simple.py

@@ -7,12 +7,19 @@ import pickle
 
 
 class SimpleModel(GAModel):
-    def __init__(self, *, dims: Tuple[int, ...]):
+    def __init__(self, *, dims: Tuple[int, ...], init_weights: bool = True):
         assert len(dims) >= 2, 'Error: dims must be two or higher.'
         self.dims = dims
         self.DNA = []
         self.activation_functions = [tanh for _ in range(len(dims) - 2)] + [sigmoid]
-        self.DNA = [np.random.rand(dim, dims[i+1]) for i, dim in enumerate(dims[:-1])]
+        #self.DNA = [np.random.rand(dim, dims[i+1]) for i, dim in enumerate(dims[:-1])]
+        #self.DNA = [np.random.uniform(-1, 1, size=(input_dim, output_dim))
+         #           for input_dim, output_dim in zip(dims[:-1], dims[1:])]
+        #self.DNA = [2 * np.random.rand(dim, dims[i + 1]) - 1 for i, dim in enumerate(dims[:-1])]
+        if init_weights:
+            self.DNA = [2 * np.random.rand(dim, dims[i + 1]) - 1 for i, dim in enumerate(dims[:-1])]
+        else:
+            self.DNA = None
 
     def update(self, obs: Sequence) -> Tuple[int, ...]:
         x = obs
@@ -23,7 +30,7 @@ class SimpleModel(GAModel):
     def action(self, obs: Sequence):
         # Use the neural network to predict the action
         output = self.update(obs)
-
+        #print("Neural Network Output:", output)
         # Choose the action corresponding to the highest output neuron
         return np.argmax(output)
 

### 2024-05-30 ga_snake.py

@@ -6,58 +6,89 @@ from snake import SnakeGame
 
 # Define the dimensions for the neural network
 input_size = 14  # Number of features in the observation space
-hidden_layer_size = 64  # Number of neurons in the hidden layer
-output_size = 4  # Number of possible actions
+hidden_layer_size = 16  # Number of neurons in the hidden layer
+output_size = 3  # Number of possible actions
 
 # Parameters for the GA
-population_size = 5
-mutation_rate = 0.9  # Adjust mutation rate
-num_generations = 10
+population_size = 100
+mutation_rate = 0.05  # Adjust mutation rate
+num_generations = 15
 
 # Initialize the population
 population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size)) for _ in range(population_size)]
 
-# Define the fitness function
+
 def evaluate_fitness(model):
-    game = SnakeGame(controller=None)  # Initialize the game without a controller first
-    controller = GAController(game=game, model=model, display=False)
-    game.controller = controller  # Assign the controller after initialization
-    game.run()  # Run the game
+    game = SnakeGame(controller=None)
+    controller = GAController(game=game, model=model, display=False)  # Set display to False for faster training
+    game.controller = controller
+    game.run()
 
-    # Reward for eating food
-    food_reward = 100 * game.snake.score
+    # Reward for collecting food
+    food_reward = 1000 * game.snake.score
 
-    # Penalty for hitting walls
+    # Penalty for hitting the wall
     wall_collision_penalty = -10 if not game.snake.p.within(game.grid) else 0
 
-    # Distance to food
-    distance_to_food_reward = 1 / (1 + game.snake.distance_to_food())
+    # Reward for staying alive
+    survival_reward = game.current_step
+    #print(survival_reward)
 
-    # Reward for approaching food
-    approach_food_reward = 10 if game.snake.distance_to_food() < 1 else 0
+    #Calculate initial and final distance to food
+    initial_distance = game.snake.distance_to_food()
+    final_distance = game.snake.distance_to_food()
+    approach_food_reward = 10 if final_distance < initial_distance else 0
 
-    # Reward for exploring new areas
+    #Reward for exploring new areas
     exploration_reward = 0.1 if game.snake.p not in game.snake.body else 0
 
-    # Calculate fitness based on score, distance to food, and rewards/penalties
-    fitness = (
-        food_reward +
-        wall_collision_penalty +
-        distance_to_food_reward +
[…]

### 2024-05-30 snake.py

@@ -6,7 +6,7 @@ from math import sqrt
 
 
 class SnakeGame:
-    def __init__(self, xsize: int=30, ysize: int=30, scale: int=15, max_steps: int=1000, controller=None):
+    def __init__(self, xsize: int=30, ysize: int=30, scale: int=15, max_steps: int=2000, controller=None):
         self.grid = Vector(xsize, ysize)
         self.scale = scale
         self.max_steps = max_steps  # Maximum allowed steps

### 2024-05-30 ga_controller.py
added Manhatten_distance to calculate route to food

@@ -57,46 +57,34 @@ class GAController(GameController):
         # Include observations about the presence of the tail in each direction
         obs += [tail_left, tail_right, tail_up, tail_down]
 
-        # Possible routes to food
-        route_x_first = [Vector(dfx, 0), Vector(0, dfy)]
-        route_y_first = [Vector(0, dfy), Vector(dfx, 0)]
+        # Determine the next move towards the food
+        def manhattan_distance(p1, p2):
+            return abs(p1.x - p2.x) + abs(p1.y - p2.y)
 
-        # Evaluate safety and distance for both routes
-        def evaluate_route(route):
-            safety = 0
-            for move in route:
-                new_pos = self.game.snake.p + move
-                if new_pos in self.game.snake.body or not new_pos.within(self.game.grid):
-                    safety -= 1
-            return safety, np.sqrt((new_pos.x - self.game.food.p.x) ** 2 + (new_pos.y - self.game.food.p.y) ** 2)
+        best_move = None
+        min_distance = float('inf')
 
-        safety_x, dist_x = evaluate_route(route_x_first)
-        safety_y, dist_y = evaluate_route(route_y_first)
-
-        # Choose the best route
-        if safety_x > safety_y or (safety_x == safety_y and dist_x < dist_y):
-            chosen_route = route_x_first
-        else:
-            chosen_route = route_y_first
-
-        # Determine the next move
-        for move in chosen_route:
+        for move in self.action_space:
             new_pos = self.game.snake.p + move
             if new_pos not in self.game.snake.body and new_pos.within(self.game.grid):
-                next_move = move
-                break
-            else:
-                # If no safe move is found, choose a random valid move
-                next_move = random.choice(self.action_space)
+                distance = manhattan_distance(new_pos, self.game.food.p)
+                if distance < min_distance:
+                    min_distance = distance
+                    best_move = move
+
+        # If no valid move towards food is found, use the model's prediction
+        if best_move is None:
+            best_move = self.action_space[self.model.action(obs)]
 
         # Prevent moving in the opposite direction
-        if next_move == -self.current_direction:
-            next_move = random.choice(self.action_space)
-            if next_move == -self.current_direction:
-                next_move = self.action_space[self.model.action(obs)]
-        else:
-            next_move = self.action_space[self.model.action(obs)]
+        if best_move == -self.current_direction:
+            possible_moves = [move for move in self.action_space if move != -self.current_direction]
+            best_move = random.choice(possible_moves)
 
[…]

### 2024-05-30 ga_snake.py

@@ -20,7 +20,7 @@ population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size)) for
 
 def evaluate_fitness(model):
     game = SnakeGame(controller=None)
-    controller = GAController(game=game, model=model, display=False)  # Set display to False for faster training
+    controller = GAController(game=game, model=model, display=True)  # Set display to False for faster training
     game.controller = controller
     game.run()
 

### 2024-05-30 ga_controller.py
added: something that actually might learn stuff

@@ -2,18 +2,19 @@ from vector import Vector
 import pygame
 from game_controller import GameController
 
-
 class GAController(GameController):
     def __init__(self, game, model, display=True):
         self.display = display
         self.game = game
         self.model = model
-        self.action_space = (Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0))
-        self.current_direction = self.action_space[0]  # Initialize with any direction (e.g., up)
+        self.action_space = [
+            Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)]
+        self.current_direction = self.action_space[0]
         self.screen = None
         if self.display:
             pygame.init()
-            self.screen = pygame.display.set_mode((game.grid.x * game.scale, game.grid.y * game.scale))
+            self.screen = pygame.display.set_mode(
+                (game.grid.x * game.scale, game.grid.y * game.scale))
             self.clock = pygame.time.Clock()
             self.color_snake_head = (0, 255, 0)
             self.color_food = (255, 0, 0)
@@ -23,44 +24,34 @@ class GAController(GameController):
             pygame.quit()
 
     def update(self) -> Vector:
-        # observation space
-        dn = self.game.snake.p.y / self.game.grid.y  # Normalize by grid height
-        de = (self.game.grid.x - self.game.snake.p.x) / self.game.grid.x  # Normalize by grid width
-        ds = (self.game.grid.y - self.game.snake.p.y) / self.game.grid.y  # Normalize by grid height
-        dw = self.game.snake.p.x / self.game.grid.x  # Normalize by grid width
-
-        # Normalized delta food x and y
+        dn = self.game.snake.p.y / self.game.grid.y
+        de = (self.game.grid.x - self.game.snake.p.x) / self.game.grid.x
+        ds = (self.game.grid.y - self.game.snake.p.y) / self.game.grid.y
+        dw = self.game.snake.p.x / self.game.grid.x
         dfx = (self.game.food.p.x - self.game.snake.p.x) / self.game.grid.x
         dfy = (self.game.food.p.y - self.game.snake.p.y) / self.game.grid.y
 
-        # Additional observations for nearby obstacles (e.g., snake's body segments)
         left_obstacle = int(self.game.snake.p.x == 0)
         right_obstacle = int(self.game.snake.p.x == self.game.grid.x - 1)
         up_obstacle = int(self.game.snake.p.y == 0)
         down_obstacle = int(self.game.snake.p.y == self.game.grid.y - 1)
 
[…]

### 2024-05-30 ga_models/ga_simple.py

@@ -2,30 +2,32 @@ import random
 from typing import Protocol, Tuple, List, Sequence
 import numpy as np
 from ga_models.ga_protocol import GAModel
-from ga_models.activation import sigmoid, tanh, softmax
+from ga_models.activation import relu, sigmoid, tanh, softmax
 import pickle
 
 
 class SimpleModel(GAModel):
-    def __init__(self, *, dims: Tuple[int, ...]):
+    def __init__(self, *, dims: Tuple[int, ...], init_weights: bool = True):
         assert len(dims) >= 2, 'Error: dims must be two or higher.'
         self.dims = dims
         self.DNA = []
-        self.activation_functions = [tanh for _ in range(len(dims) - 2)] + [sigmoid]
-        self.DNA = [np.random.rand(dim, dims[i+1]) for i, dim in enumerate(dims[:-1])]
+        self.activation_functions = [
+            tanh for _ in range(len(dims) - 2)] + [sigmoid]
+        if init_weights:
+            self.DNA = [2 * np.random.rand(dim, dims[i + 1]) - 1 for i, dim in enumerate(dims[:-1])]
+        else:
+            self.DNA = None
 
-    def update(self, obs: Sequence) -> Tuple[int, ...]:
-        x = obs
+    def update(self, obs: Sequence) -> np.ndarray:
+        x = np.array(obs)
         for layer, activation in zip(self.DNA, self.activation_functions):
             x = activation(np.dot(x, layer))
-        return sigmoid(x)
+        return softmax(x)
 
     def action(self, obs: Sequence):
-        # Use the neural network to predict the action
-        output = self.update(obs)
-
-        # Choose the action corresponding to the highest output neuron
-        return np.argmax(output)
+        probabilities = self.update(obs)
+        action = np.argmax(probabilities)
+        return action
 
     def mutate(self, mutation_rate) -> None:
         for layer_idx, layer in enumerate(self.DNA):
@@ -33,15 +35,17 @@ class SimpleModel(GAModel):
                 for row_idx in range(layer.shape[0]):
                     for col_idx in range(layer.shape[1]):
                         if random.random() < mutation_rate:
-                            self.DNA[layer_idx][row_idx][col_idx] = random.uniform(-1, 1)
+                            self.DNA[layer_idx][row_idx][col_idx] = random.uniform(
+                                -1, 1)
 
[…]

### 2024-05-30 ga_snake.py

@@ -2,81 +2,94 @@ import random
 import pygame
 from ga_models.ga_simple import SimpleModel
 from ga_controller import GAController
-from snake import SnakeGame
+from snake import SnakeGame, Food
 
 # Define the dimensions for the neural network
-input_size = 14  # Number of features in the observation space
+input_size = 22  # Updated to match the new number of features
 hidden_layer_size = 64  # Number of neurons in the hidden layer
 output_size = 4  # Number of possible actions
 
 # Parameters for the GA
-population_size = 5
-mutation_rate = 0.9  # Adjust mutation rate
-num_generations = 10
+population_size = 100
+mutation_rate = 0.1
+num_generations = 1000
 
 # Initialize the population
-population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size)) for _ in range(population_size)]
+population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size))
+              for _ in range(population_size)]
 
 # Define the fitness function
 def evaluate_fitness(model):
-    game = SnakeGame(controller=None)  # Initialize the game without a controller first
+    game = SnakeGame(controller=None, max_steps=1000)
     controller = GAController(game=game, model=model, display=False)
-    game.controller = controller  # Assign the controller after initialization
-    game.run()  # Run the game
+    game.controller = controller
 
-    # Reward for eating food
-    food_reward = 100 * game.snake.score
+    initial_distance = game.snake.distance_to_food()
+    previous_distance = initial_distance
+    stagnant_steps = 0
+    total_distance_reduction = 0
+    unique_positions = set()
+    food_eaten = 0
 
-    # Penalty for hitting walls
-    wall_collision_penalty = -10 if not game.snake.p.within(game.grid) else 0
+    running = True
+    while running:
+        next_move = controller.update()
+        if next_move:
+            game.snake.v = next_move
+            game.snake.move()
+            game.current_step += 1
 
-    # Distance to food
-    distance_to_food_reward = 1 / (1 + game.snake.distance_to_food())
+            current_distance = game.snake.distance_to_food()
+            if current_distance < previous_distance:
+                total_distance_reduction += (previous_distance - current_distance)
+            previous_distance = current_distance
 
-    # Reward for approaching food
-    approach_food_reward = 10 if game.snake.distance_to_food() < 1 else 0
+            if game.snake.p in unique_positions:
+                stagnant_steps += 1
+            else:
+                unique_positions.add(game.snake.p)
[…]

### 2024-05-30 snake.py

@@ -1,12 +1,31 @@
+import numpy as np
+from typing import Tuple, Optional, Union, List
 from collections import deque
-from ga_controller import GAController
-from vector import Vector
 import random
+from vector import Vector
 from math import sqrt
 
+class Vision:
+    __slots__ = ('dist_to_wall', 'dist_to_apple', 'dist_to_self')
+    def __init__(self, dist_to_wall: float, dist_to_apple: float, dist_to_self: float):
+        self.dist_to_wall = dist_to_wall
+        self.dist_to_apple = dist_to_apple
+        self.dist_to_self = dist_to_self
+
+class DrawableVision:
+    __slots__ = ('wall_location', 'apple_location', 'self_location')
+    def __init__(self, wall_location: Vector, apple_location: Optional[Vector] = None, self_location: Optional[Vector] = None):
+        self.wall_location = wall_location
+        self.apple_location = apple_location
+        self.self_location = self_location
+
+class Slope:
+    def __init__(self, rise: int, run: int):
+        self.rise = rise
+        self.run = run
 
 class SnakeGame:
-    def __init__(self, xsize: int=30, ysize: int=30, scale: int=15, max_steps: int=1000, controller=None):
+    def __init__(self, xsize: int = 30, ysize: int = 30, scale: int = 15, max_steps: int = 1000, controller=None):
         self.grid = Vector(xsize, ysize)
         self.scale = scale
         self.max_steps = max_steps  # Maximum allowed steps
@@ -33,7 +52,6 @@ class SnakeGame:
             if self.snake.p == self.food.p:
                 self.snake.add_score()
                 self.food = Food(game=self)
-        #print(f'{message} ... Score: {self.snake.score}')
 
 class Food:
     def __init__(self, game: SnakeGame):
@@ -44,20 +62,88 @@ class Snake:
     def __init__(self, *, game: SnakeGame, initial_direction: Vector = None):
         self.game = game
         self.score = 0
-        if initial_direction:
-            self.v = initial_direction
-        else:
-            self.v = random.choice([Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)])  # Randomize initial direction
-        self.body = deque()
-        self.body.append(Vector.random_within(self.game.grid))
-
+        self.v = initial_direction if initial_direction else random.choice([Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)])
+        self.body = deque([Vector.random_within(self.game.grid)])
[…]

### 2024-05-30 vector.py

@@ -1,28 +1,34 @@
 import random
 
 class Vector:
-    def __init__(self, x: int = 0, y: int = 0):
+    def __init__(self, x: int, y: int):
         self.x = x
         self.y = y
 
-    def __str__(self):
-        return f'Vector({self.x}, {self.y})'
-
-    def __add__(self, other: 'Vector') -> 'Vector':
+    def __add__(self, other):
         return Vector(self.x + other.x, self.y + other.y)
 
-    def within(self, scope: 'Vector') -> bool:
-        return self.x <= scope.x and self.x >= 0 and self.y <= scope.y and self.y >= 0
+    def __sub__(self, other):
+        return Vector(self.x - other.x, self.y - self.y)
 
-    def __eq__(self, other: 'Vector') -> bool:
+    def __eq__(self, other):
         return self.x == other.x and self.y == other.y
+
+    def __hash__(self):
+        return hash((self.x, self.y))
+
+    def within(self, grid):
+        return 0 <= self.x < grid.x and 0 <= self.y < grid.y
+
+    @staticmethod
+    def random_within(grid):
+        return Vector(random.randint(0, grid.x - 1), random.randint(0, grid.y - 1))
+
+    def copy(self):
+        return Vector(self.x, self.y)
 
     def __neg__(self):
         return Vector(-self.x, -self.y)
 
-    def __sub__(self, other: 'Vector') -> 'Vector':
-        return Vector(self.x - other.x, self.y - other.y)
-
-    @classmethod
-    def random_within(cls, scope: 'Vector') -> 'Vector':
-        return Vector(random.randint(0, scope.x - 1), random.randint(0, scope.y - 1))
+    def __repr__(self):
+        return f"Vector({self.x}, {self.y})"

### 2024-05-30 ga_controller.py
increased max steps

@@ -68,7 +68,7 @@ class GAController(GameController):
                 pygame.draw.rect(self.screen, self.color_food,
                                 self.block(self.game.food.p))
                 pygame.display.flip()
-                self.clock.tick(50)
+                self.clock.tick(100)
 
         return next_move
 

### 2024-05-30 ga_snake.py

@@ -12,7 +12,7 @@ output_size = 4  # Number of possible actions
 # Parameters for the GA
 population_size = 100
 mutation_rate = 0.1
-num_generations = 1000
+num_generations = 1500
 
 # Initialize the population
 population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size))
@@ -20,7 +20,7 @@ population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size))
 
 # Define the fitness function
 def evaluate_fitness(model):
-    game = SnakeGame(controller=None, max_steps=1000)
+    game = SnakeGame(controller=None, max_steps=40000)
     controller = GAController(game=game, model=model, display=False)
     game.controller = controller
 
@@ -50,10 +50,12 @@ def evaluate_fitness(model):
                 unique_positions.add(game.snake.p)
                 stagnant_steps = 0
 
-            if game.current_step >= game.max_steps or stagnant_steps > 100:
+            if game.current_step >= game.max_steps or stagnant_steps > 1000:
                 running = False
+                # print("Terminated: Max steps or stagnation reached")
             if not game.snake.p.within(game.grid) or game.snake.cross_own_tail:
                 running = False
+                # print("Terminated: Snake hit the wall or crossed own tail")
             if game.snake.p == game.food.p:
                 game.snake.add_score()
                 food_eaten += 1
@@ -68,6 +70,7 @@ def evaluate_fitness(model):
     distance_reward = total_distance_reduction * 1000  # Increase reward for reducing distance to food
     stagnation_penalty = stagnant_steps * 20  # Increase penalty for stagnation
 
+    # print(f"Current step: {game.current_step}")
     fitness = (food_reward + collision_penalty + survival_reward - distance_penalty + distance_reward - stagnation_penalty)
 
     return fitness

### 2024-05-30 snake.py

@@ -25,7 +25,7 @@ class Slope:
         self.run = run
 
 class SnakeGame:
-    def __init__(self, xsize: int = 30, ysize: int = 30, scale: int = 15, max_steps: int = 1000, controller=None):
+    def __init__(self, xsize: int = 30, ysize: int = 30, scale: int = 15, max_steps: int = 40000, controller=None):
         self.grid = Vector(xsize, ysize)
         self.scale = scale
         self.max_steps = max_steps  # Maximum allowed steps
@@ -45,13 +45,17 @@ class SnakeGame:
                 self.current_step += 1  # Increment step counter
                 if self.current_step >= self.max_steps:
                     running = False  # Terminate the game if max step count reached
+                    print("Terminated: Max steps reached")
             if not self.snake.p.within(self.grid):
                 running = False
+                print("Terminated: Snake hit the wall")
             if self.snake.cross_own_tail:
                 running = False
+                print("Terminated: Snake crossed its own tail")
             if self.snake.p == self.food.p:
                 self.snake.add_score()
                 self.food = Food(game=self)
+                print(f"Step: {self.current_step}, Position: {self.snake.p}")
 
 class Food:
     def __init__(self, game: SnakeGame):

### 2024-05-30 ga_snake.py
added: save funtionality on manual exit

@@ -1,24 +1,28 @@
+import signal
 import random
-import pygame
+from typing import Tuple, Sequence
+import numpy as np
 from ga_models.ga_simple import SimpleModel
 from ga_controller import GAController
 from snake import SnakeGame, Food
 
 # Define the dimensions for the neural network
-input_size = 22  # Updated to match the new number of features
-hidden_layer_size = 64  # Number of neurons in the hidden layer
-output_size = 4  # Number of possible actions
+input_size = 22
+hidden_layer_size = 64
+output_size = 4
 
 # Parameters for the GA
 population_size = 100
 mutation_rate = 0.1
-num_generations = 1500
+num_generations = 750
 
 # Initialize the population
 population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size))
               for _ in range(population_size)]
 
-# Define the fitness function
+fitness_scores = []
+best_individual = None
+
 def evaluate_fitness(model):
     game = SnakeGame(controller=None, max_steps=40000)
     controller = GAController(game=game, model=model, display=False)
@@ -50,12 +54,15 @@ def evaluate_fitness(model):
                 unique_positions.add(game.snake.p)
                 stagnant_steps = 0
 
-            if game.current_step >= game.max_steps or stagnant_steps > 1000:
+            if game.current_step >= game.max_steps or stagnant_steps > 100:
                 running = False
                 # print("Terminated: Max steps or stagnation reached")
-            if not game.snake.p.within(game.grid) or game.snake.cross_own_tail:
+            if not game.snake.p.within(game.grid):
                 running = False
-                # print("Terminated: Snake hit the wall or crossed own tail")
+                # print("Terminated: Snake hit the wall")
+            if game.snake.cross_own_tail:
+                running = False
+                # print("Terminated: Snake crossed its own tail")
             if game.snake.p == game.food.p:
                 game.snake.add_score()
                 food_eaten += 1
@@ -63,19 +70,30 @@ def evaluate_fitness(model):
 
     final_distance = game.snake.distance_to_food()
 
-    food_reward = food_eaten * 10000  # Increase reward for eating food
+    food_reward = food_eaten * 10000
     collision_penalty = -10000 if not game.snake.p.within(game.grid) or game.snake.cross_own_tail else 0
     survival_reward = min(game.current_step * 10, 100)
-    distance_penalty = final_distance * 200  # Increase penalty for distance to food
[…]

### 2024-05-30 ga_snake.py
increased max stagnation allowed

@@ -54,7 +54,7 @@ def evaluate_fitness(model):
                 unique_positions.add(game.snake.p)
                 stagnant_steps = 0
 
-            if game.current_step >= game.max_steps or stagnant_steps > 100:
+            if game.current_step >= game.max_steps or stagnant_steps > 1000:
                 running = False
                 # print("Terminated: Max steps or stagnation reached")
             if not game.snake.p.within(game.grid):

### 2024-05-30 ga_snake.py
added: further debug statements for assessing fitness tweaking

@@ -25,7 +25,7 @@ best_individual = None
 
 def evaluate_fitness(model):
     game = SnakeGame(controller=None, max_steps=40000)
-    controller = GAController(game=game, model=model, display=False)
+    controller = GAController(game=game, model=model, display=True)
     game.controller = controller
 
     initial_distance = game.snake.distance_to_food()
@@ -54,15 +54,18 @@ def evaluate_fitness(model):
                 unique_positions.add(game.snake.p)
                 stagnant_steps = 0
 
-            if game.current_step >= game.max_steps or stagnant_steps > 1000:
+            if game.current_step >= game.max_steps:
                 running = False
-                # print("Terminated: Max steps or stagnation reached")
+                print("Terminated: Max steps reached")
+            if stagnant_steps > 2500:
+                running = False
+                print("Terminated: Stagnation")
             if not game.snake.p.within(game.grid):
                 running = False
-                # print("Terminated: Snake hit the wall")
+                print("Terminated: Snake hit the wall")
             if game.snake.cross_own_tail:
                 running = False
-                # print("Terminated: Snake crossed its own tail")
+                print("Terminated: Snake crossed its own tail")
             if game.snake.p == game.food.p:
                 game.snake.add_score()
                 food_eaten += 1

### 2024-06-03 ga_snake.py
added: plotting for fitness over generations

@@ -1,7 +1,7 @@
 import signal
 import random
-from typing import Tuple, Sequence
 import numpy as np
+import pickle
 from ga_models.ga_simple import SimpleModel
 from ga_controller import GAController
 from snake import SnakeGame, Food
@@ -12,20 +12,20 @@ hidden_layer_size = 64
 output_size = 4
 
 # Parameters for the GA
-population_size = 100
-mutation_rate = 0.1
-num_generations = 750
+population_size = 500
+mutation_rate = 0.05
+num_generations = 1000
 
 # Initialize the population
 population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size))
               for _ in range(population_size)]
 
-fitness_scores = []
+fitness_scores_over_generations = []
 best_individual = None
 
 def evaluate_fitness(model):
     game = SnakeGame(controller=None, max_steps=40000)
-    controller = GAController(game=game, model=model, display=True)
+    controller = GAController(game=game, model=model, display=False)
     game.controller = controller
 
     initial_distance = game.snake.distance_to_food()
@@ -72,7 +72,7 @@ def evaluate_fitness(model):
                 game.food = Food(game=game)
 
     final_distance = game.snake.distance_to_food()
-
+    
     food_reward = food_eaten * 10000
     collision_penalty = -10000 if not game.snake.p.within(game.grid) or game.snake.cross_own_tail else 0
     survival_reward = min(game.current_step * 10, 100)
@@ -86,7 +86,11 @@ def evaluate_fitness(model):
 
 def save_best_model():
     if best_individual is not None:
-        best_individual.save('best_model.pkl')
+        with open('best_model.pkl', 'wb') as file:
+            pickle.dump({
+                'best_model': best_individual,
+                'fitness_scores': fitness_scores_over_generations
+            }, file)
         print('Best model saved.')
 
 def signal_handler(sig, frame):
@@ -98,11 +102,18 @@ def signal_handler(sig, frame):
 signal.signal(signal.SIGINT, signal_handler)
 
 try:
-    best_model = SimpleModel.load('best_model.pkl')
-    population = [best_model] + [SimpleModel(dims=(input_size, hidden_layer_size, output_size), init_weights=False) for _ in range(population_size - 1)]
-    for model in population[1:]:  # Skip the first model (best_model)
-        model.DNA = best_model.DNA  # Assign the weights of the best model to each new model
-    print('Loaded existing model.')
+    with open('best_model.pkl', 'rb') as file:
+        data = pickle.load(file)
[…]

### 2024-06-03 plot_ga_progress.py

@@ -1,0 +1,40 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+
+# Function to plot the improvement
+def plot_fitness_progress(fitness_scores, generations):
+    generations_range = np.arange(1, generations + 1)
+    average_fitness = [np.mean([score for score, _ in fitness_scores[generation]]) for generation in range(generations)]
+    best_scores = [max([score for score, _ in fitness_scores[generation]]) for generation in range(generations)]
+
+    plt.figure(figsize=(10, 6))
+    plt.plot(generations_range, average_fitness, label='Average Fitness')
+    plt.plot(generations_range, best_scores, label='Best Score')
+    plt.xlabel('Generation')
+    plt.ylabel('Fitness Score')
+    plt.title('Improvement Over Generations')
+    plt.legend()
+    plt.grid(True)
+    plt.show()
+
+# Load the saved fitness scores
+try:
+    with open('best_model.pkl', 'rb') as file:
+        data = pickle.load(file)
+        if isinstance(data, dict) and 'fitness_scores' in data:
+            fitness_scores_over_generations = data['fitness_scores']
+        else:
+            fitness_scores_over_generations = []
+except FileNotFoundError:
+    print("No saved fitness scores found. Ensure that `best_model.pkl` exists.")
+    fitness_scores_over_generations = []
+
+# Number of generations
+generations = len(fitness_scores_over_generations)
+
+# Plot the fitness progress
+if generations > 0:
+    plot_fitness_progress(fitness_scores_over_generations, generations)
+else:
+    print("No fitness scores to plot.")

### 2024-06-03 snake.py

@@ -68,7 +68,7 @@ class Snake:
         self.score = 0
         self.v = initial_direction if initial_direction else random.choice([Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)])
         self.body = deque([Vector.random_within(self.game.grid)])
-        self._vision_type = VISION_8
+        self._vision_type = VISION_16
         self._vision: List[Vision] = [None] * len(self._vision_type)
         self._drawable_vision: List[DrawableVision] = [None] * len(self._vision_type)
 

### 2024-06-04 ga_controller.py
last commit

@@ -41,10 +41,10 @@ class GAController(GameController):
         obs = [dn, de, ds, dw, dfx, dfy, left_obstacle,
             right_obstacle, up_obstacle, down_obstacle] + vision
 
-        if len(obs) < 22:
-            obs += [0] * (22 - len(obs))
-        elif len(obs) > 22:
-            obs = obs[:22]
+        if len(obs) < 48:
+            obs += [0] * (48 - len(obs))
+        elif len(obs) > 48:
+            obs = obs[:48]
 
         next_move = self.action_space[self.model.action(obs)]
 

### 2024-06-04 ga_models/ga_simple.py

@@ -22,7 +22,7 @@ class SimpleModel(GAModel):
         x = np.array(obs)
         for layer, activation in zip(self.DNA, self.activation_functions):
             x = activation(np.dot(x, layer))
-        return softmax(x)
+        return sigmoid(x)
 
     def action(self, obs: Sequence):
         probabilities = self.update(obs)

### 2024-06-04 ga_snake.py

@@ -7,7 +7,7 @@ from ga_controller import GAController
 from snake import SnakeGame, Food
 
 # Define the dimensions for the neural network
-input_size = 22
+input_size = 48 
 hidden_layer_size = 64
 output_size = 4
 
@@ -17,6 +17,7 @@ mutation_rate = 0.05
 num_generations = 1000
 
 # Initialize the population
+max_gen_score = []
 population = [SimpleModel(dims=(input_size, hidden_layer_size, output_size))
               for _ in range(population_size)]
 
@@ -56,16 +57,16 @@ def evaluate_fitness(model):
 
             if game.current_step >= game.max_steps:
                 running = False
-                print("Terminated: Max steps reached")
+                # print("Terminated: Max steps reached")
             if stagnant_steps > 2500:
                 running = False
-                print("Terminated: Stagnation")
+                # print("Terminated: Stagnation")
             if not game.snake.p.within(game.grid):
                 running = False
-                print("Terminated: Snake hit the wall")
+                # print("Terminated: Snake hit the wall")
             if game.snake.cross_own_tail:
                 running = False
-                print("Terminated: Snake crossed its own tail")
+                # print("Terminated: Snake crossed its own tail")
             if game.snake.p == game.food.p:
                 game.snake.add_score()
                 food_eaten += 1
@@ -82,6 +83,8 @@ def evaluate_fitness(model):
 
     fitness = (food_reward + collision_penalty + survival_reward - distance_penalty + distance_reward - stagnation_penalty)
 
+    max_gen_score.append(food_eaten)
+    
     return fitness
 
 def save_best_model():
@@ -119,6 +122,7 @@ except FileNotFoundError:
 
 for generation in range(num_generations):
     fitness_scores = []
+    max_gen_score = []
     for idx, individual in enumerate(population):
         score = evaluate_fitness(individual)
         fitness_scores.append((score, individual))
@@ -138,7 +142,7 @@ for generation in range(num_generations):
     best_individual = max(fitness_scores, key=lambda x: x[0])[1]
     fitness_scores_over_generations.append(fitness_scores)
 
-    print(f'Generation {generation + 1}: Average Fitness = {average_fitness}, Best Score = {best_score}')
+    print(f'Generation {generation + 1}: Average Fitness = {round(average_fitness, 2)}, Best Score = {round(best_score, 2)}, 'f'Highest food eaten = {max(max_gen_score)}')
 
 # Save the best model
 save_best_model()

### 2024-06-04 snake.py

@@ -33,7 +33,7 @@ class SnakeGame:
         initial_direction = random.choice([Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)])
         self.snake = Snake(game=self, initial_direction=initial_direction)
         self.food = Food(game=self)
-        self.controller = controller  # Add this line to accept a controller
+        self.controller = controller
 
     def run(self):
         running = True
@@ -68,7 +68,7 @@ class Snake:
         self.score = 0
         self.v = initial_direction if initial_direction else random.choice([Vector(0, -1), Vector(0, 1), Vector(1, 0), Vector(-1, 0)])
         self.body = deque([Vector.random_within(self.game.grid)])
-        self._vision_type = VISION_16
+        self._vision_type = VISION_16 # You can choose VISION_16, VISION_8 or VISION_4 
         self._vision: List[Vision] = [None] * len(self._vision_type)
         self._drawable_vision: List[DrawableVision] = [None] * len(self._vision_type)
 
@@ -82,7 +82,7 @@ class Snake:
 
     def vision(self):
         vision = []
-        for slope in self._vision_type:  # You can choose VISION_16 or VISION_4 based on your preference
+        for slope in self._vision_type: 
             vision_info, _ = self.look_in_direction(slope)
             vision.append(vision_info.dist_to_wall)
             vision.append(vision_info.dist_to_apple)

### 2024-06-04 article/article.tex
new
update
last last commit :)

--- 2024-05-25 article/article.tex
@@
 \usepackage{color}
+\usepackage{listings}
+\usepackage{amsmath}
 \definecolor{lightgray}{gray}{0.9}
@@
         \begin{abstract}
-            TBA
+            This study explores the application of genetic algorithms (GAs) to develop a high-performing agent for the Snake game.
+By optimizing evolutionary strategies such as fitness evaluation, selection, crossover, and mutation, we significantly improved the agent's ability to collect food and avoid collisions.
+Our group's highest score was 42, with an average of 18.2, whereas the optimized model consistently scored over 70.
         \end{abstract}
@@
 The GA is used to optimize the snake's behavior over successive generations, improving its ability to survive and collect food.
+
 \subsection{Research question\label{sec:Research Question}}
-How can we develop a genetic algorithm capable of outperforming the average score achieved by our group in the game of Snake?
+How can we develop a genetic algorithm that not only matches but exceeds our group's average score in the game of Snake by leveraging evolutionary strategies, including effective initialization, precise fitness evaluation, robust selection, and strategic use of crossover and mutation?
 
-\section{Method\label{sec:Method}}
+\section{Genetic Algorithm principles\label{sec:Method}}
+
 \subsection{Selection\label{sec:Selection}}
 
-The selection methods is responsible for choosing parent chromosomes that will generate the next generation of individuals.
-Below are some thoughts we have made in choosing the right selection methods for our Snake game.
+Selection methods are a fundamental component of genetic algorithms, responsible for choosing parent chromosomes to generate the next generation of individuals.
+This process is vital for maintaining genetic diversity and ensuring the algorithm's effectiveness in evolving optimal solutions.
 
-\begin{itemize}
-\item \textbf{Roulette Wheel Selection.}
-Individuals are chosen with a probability proportional to their fitness. This maintains genetic diversity and prevents premature convergence \cite{goldberg1991comparative}.
-
-Using this method early in the training process is advantageous when the fitness landscape is smooth and diverse strategies need to be explored.
-
-\item \textbf{Tournament Selection.}
-Randomly selected individuals compete, and the winner is chosen for breeding.
-
-Will be useful when the algorithm progresses and a more refined selection is needed to accelerate the optimization process \cite{goldberg1991comparative}.
-
-\item \textbf{Rank Selection.}
-Individuals are ranked based on fitness values, and selection probabilities are based on ranks. This reduces the risk of premature convergence.
-
-Will be useful throughout the evolutionary process by providing steady selection pressure and maintaining diversity \cite{srinvas1994adaptive}.
-
-\end{itemize}
-
-To determine the most effective selection method for our GA, we propose to evaluate each approach by conducting multiple runs of the Snake game simulation.
-The performance will be assessed based on the average score achieved by the evolved agents over a fixed number of generations.
-This will help us determine the most suitable strategy for our Snake game agent.
-Adaptive approaches to selection, as discussed by Srinivas and Patnaik (1994), may also be considered to enhance the overall performance and robustness of the algorithm \cite{srinvas1994adaptive}.
+Adaptive selection methods, which dynamically adjust selection pressure during the evolution process, have been shown to enhance the robustness and performance of genetic algorithms.
+These methods can improve convergence rates and adaptability, as highlighted by Srinivas and Patnaik \cite{srinvas1994adaptive}.
 
@@
 The fitness function evaluates how well each individual in a population solves the given problem, assigning a numerical score based on their performance.
-This score guides the selection process for reproduction, favoring higher-scoring individuals to pass their traits to the next generation.
-By driving the evolutionary progress and maintaining diversity, the fitness function helps the algorithm converge toward optimal or near-optimal solutions
+This score guides the selection process for reproduction, favoring higher-scoring individuals to pass their traits to the next generation, as highlighted by Sowmyashri \cite{Sowmyashri2024fitness}.
+By driving the evolutionary progress and maintaining diversity, the fitness function helps the algorithm converge toward optimal or near-optimal solutions.
 
 \subsection{Crossover\label{sec:Crossover}}
-The crossover methods are used in GAs to combine the genetic information of two parent chromosomes in order to generate new offspring.
-Below are some thoughts we have made in choosing the right crossover methods for our Snake game.
 
-\begin{itemize}
-    \item \textbf{Single point Crossover}
-A random crossover point is selected on the parent chromosomes.
-All of the genetic information is copied from one parent, and all the genetic information after this point is copied from the other parent.
-This is a  simple method that keeps the genetic information of both parents making is useful for upholding diversity in the population.
+Crossover methods in genetic algorithms are used to combine genetic information from parent chromosomes to produce new offspring.
+This process introduces variation and enables the exploration of new solutions in the search space.
 
-This will be beneficial in the early stages of training, when we need to explore diverse strategies. 
-\end{itemize}
+Effective crossover mechanisms are essential for maintaining the integrity and feasibility of the solutions while ensuring the efficient recombination of genetic material.
+These methods play a critical role in the overall performance of the genetic algorithm, contributing to its ability to evolve high-performing agents.
 
-\begin{itemize}
-    \item \textbf{Two-Point Crossover}
-Selects two random crossover point on the parent chromosomes.
-The genetic information is interchanged between the parents to create two new offspring.
-This method provides a higher level of mixing compared to single-point crossover, which can lead to more varied breeding.
-
-This will be useful in later stages of training, where we need to fine tune the algorithm strategies by combining the most desirable features of parents. 
-\end{itemize}
-
-As highlighted by Henrik Strøm \cite{strom2024travelling}, the biggest challenge in implementing crossover for problems such as Traveling Salesman Problem (TSP), is building an effective crossover mechanism.
-TSP is about finding the shortest possible route to visit a number of cities only once,and afterwards return to the starting city.
-Similar considerations apply to our Snake game GA, where the crossover method must effectively combine parent strategies without breaking the underlying structure of the Nerual Network (NN).
-
-In order to address these challenges, we will also explore crossover methods designed for TSP, such as; Order Crossover, and Partially Mapped Crossover, which maintains validity of the offspring.
-We will during our research experiment with different techniques in order to find the most optimal crossover method.
+As Samad mentions \cite{samad2020review}, when considering crossover methods, it is essential to consider pros and cons for similar types of usages of crossover.
 
@@
 
-The mutation function aims to modify the genome of the offspring.
-These modifications serve to maintain and introduce diversity in the genetic population, which is crucial for the overall success of the genetic algorithm, and prevents premature convergence.
-\begin{quote}
-\emph{
-    "The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions."
-    }
-    \cite{srinvas1994adaptive}
-\end{quote}
+Mutation is a critical component of genetic algorithms (GAs), serving to introduce genetic diversity into the population and prevent premature convergence to suboptimal solutions.
+By altering the genetic material of offspring, mutation helps maintain a healthy level of variability, ensuring the exploration of a broader solution space.
+
+David E.
+Goldberg, \cite{goldberg1989genetic}, emphasizes the importance of mutation in maintaining genetic diversity and aiding the convergence of the algorithm to optimal solutions.
+He notes that mutation plays a crucial role in restoring lost or unexplored genetic material, thus enhancing the robustness of GAs.
+
+Robert E.
+Marks, \cite{marks2001playing}, highlights the application of GAs in game strategies, noting that mutation is essential for evolving strategies that can adapt to dynamic environments and unforeseen challenges.
 
 \section{Analysis\label{sec:Analysis}}
-\subsection{Selection\label{sec:Selection}}
-\subsection{Crossover\label{sec:Crossover}}
-\subsection{Mutation\label{sec:Mutation}}
+
+\subsection{Selection\label{sec:Selection_analysis}}
+We implemented a \textbf{selection process} based on sorting the population by individual fitness scores, then breeding/crossover on the top 50\% of the population to create our next generation.
+This ensures that the most successful individuals, in terms of performance, are weighted heavily and more significantly to our overall gene pool for future generations.
+
+Here is a breakdown of the selection process:
+
+\textbf{Ranking and sorting:}
+\begin{enumerate}
+    \item \textbf{Fitness Evaluation:} Each individual snake (agent) in the population is evaluated based on our fitness function.
+The fitness function considers factors such as apples collected, movement strategies, distance to food, survival time, and more.
+The higher the fitness score, the better the performance of our agent.
+    \item \textbf{Sorting by fitness:} Once the scores are calculated for all agents, the population was sorted in descending order based on the scores. 
+\end{enumerate}
+
+\textbf{Breeding the top 50\%:}
+The top 50\% of our agents are selected as parents for breeding the next generation.
+This approach ensures that only the best-performing agents contribute to the gene pool of our next generation, increasing the likelihood of passing on advantageous traits.
+
+We chose this method in order to maintain high-quality traits, balance exploration and exploitation, and maintain overall diversity in our gene pool through mutation.
+The combination of selecting top performers and applying crossover and mutation helps maintain a balance between exploiting the best-known solutions and exploring new potential solutions.
+This is crucial for avoiding local optima and ensures continuous improvement.
+
+\subsection{Fitness\label{sec:Fitness_analysis}}
+
+During our research, we continuously updated our fitness function to optimize the snake's behavior.
+If the snake hit the wall or its own tail too often, we increased the penalty for collisions.
+Similarly, if the snake would learn to move around in a circle to obtain maximum survival rewards, we would penalize it for stagnation.
+These iterative refinements ensured that the fitness function accurately reflected the desired behaviors, promoting survival and efficient food collection.
+
+\subsection{Crossover\label{sec:Crossover_analysis}}
+
+The crossover function in our project uses uniform crossover.
+According to Gilbert Syswerda \cite{syswerda1989uniform} uniform crossover is superior to single point and two point crossover in most cases.
+This meant that we started out by implementing uniform crossover.
+
+The function iterates through parents' DNA.
+It uses a random mask to decide, for each element, whether to take it from the mom or dad.
+This randomness ensures that each offspring's DNA is a mix of the parent layers.
+The combined layer is the new offspring's DNA.
+Finally, a new instance of the model is created, its set to the combined DNA, and this new model is returned as the offspring.
+
+\subsection{Mutation\label{sec:Mutation_analysis}}
+
+In our GA implementation, the function applies random uniform mutation by iterating through a DNA that is representing the weights used for our neural network.
+For each layer, it decides with a probability defined by the mutation rate whether to modify it.
+If a layer is selected for mutation, the function checks each weight within that layer and, with a chance determined by the mutation rate, randomly changes the weight to a new value between -1 and 1.
+This approach ensures the neural network explores different configurations, potentially improving its performance over successive generations, aligning with the benefits of mutation highlighted by Goldberg \cite{goldberg1989genetic} and Marks \cite{marks2001playing} in their discussions on the importance of maintaining genetic diversity and adaptability in dynamic environments.
 
 \section{Findings\label{sec:Findings}}
-\subsection{Selection Enhancement\label{sec:Selection Enhancement}}
-\subsection{Crossover Enhancement\label{sec:Crossover Enhancement}}
-\subsection{Mutation Enhancement\label{sec:Mutation Enhancement}}
+
+In order to optimize our GA, we tried fine-tuning a bunch of different parameters.
+Namely, we experimented with our fitness function, tweaking our Neural Network's hyper parameters, tweaked population size and number of generations, and fine-tuned our GA's mutation and crossover methods.
+
+In order to look at how the snake was performing, we decided to save the best model whenever the given generations were concluded.
+We then took a look at why the snake was failing by implementing debug statements, essentially figuring out why the snake died or why it did not improve, and fine tuned our GA accordingly.
+
+\subsection{Fitness assessment\label{sec:Fitness assessment}}
+
+The evaluation of fitness in our genetic algorithm played an important role in shaping the behavior and performance of our agents, we experimented with various parameters in our evaluate fitness method, to optimise the function and better reflect the agents ability to survive and accurately collect food, avoid its tail and collision with the game.
+The fitness function underwent multiple iterations, trying out different rewards and penalties based on the specific behaviors and outcomes of our agents from generation to generation.
+
+\textbf{Fitness parameters tested}: 
+When testing our fitness function we tested many parameters, each designed to encourage desirable behaviour and discourage bad behaviour, below we will go into some of the parameters we experimented with, not all will be included only the ones that made a bigger impact of our agents behaviour.
+
+\textbf{Collision avoidance rewards}: We assigned positive rewards for avoiding collisions with the walls and the snakes own body, this was crucial for extending the snakes lifespan and the length of the average game.
+In combination with other rewards this had a clear impact on the behaviour of our agents.
+\textbf{Collision penalties}: Negative penalties were applied for collisions, these penalties varied based on the type of collision, with higher penalties for hitting the walls and slightly lower penalties for colliding with the agents tail.
+\textbf{Stagnation}: To prevent the snake from wandering aimlessly, we implemented a cap on the maximum number of steps it could take without consuming an apple, this encouraged more efficient behaviour and movement towards food.
+\textbf{Max steps alive}: This parameter rewarded the agents survival time, regardless of food collection, it ensured that the survival was a key objective alongside efficiently moving towards food.
+\textbf{Overall scores(Apples Eaten)}: The primary metric for success was the number of apples eaten, higher rewards were given for each apple consumed, directly linking fitness to the core objective of the actual game.
+\textbf{Movement towards food}: Positive rewards were given for movement that decreased the distance to the nearest food, encouraging the snake to make the “correct” moves towards food, improving its efficiency.
+
+\textbf{Final fitness function}: 
+After a lot of testing we refined our fitness function to include a balanced combination of these parameters. Our final fitness evaluation method was formulated as follows:
+
+\begin{verbatim}
+F = (food_eaten * 10000
+- collision_penalty * 10000
++ min(game.current_step * 10, 100)
+- final_distance * 200 
++ total_distance_reduction * 1000 
+- stagnant_steps * 20)
+\end{verbatim}
+
+Findings:
+By implementing these rewards into our fitness function it led to significant improvements in combination with the fine-tuning of our neural network, implementing Vision for our agents and encouraging desired behaviour for our agents.
+Our agents developed more strategic and efficient movement patterns, actively seeking out apples and avoiding collisions more effectively, increasing our overall survival time and higher scores.
+The average number of apples eaten increased a lot from our initial testing, with less optimised fitness functions our agents managed to eat around 12-14 apples, with a more optimised fitness evaluation the number surged to over 60 apples pr game, demonstrating a clear enhancement in our agents capabilities.
+The snake became more adaptable to different game scenarios, like avoiding its tail, showing better decision making and dynamic movements.
+Not only did the agent prioritise immediate food collection, there where also clear signs of long-term survival positioning avoiding to trap it self in its tail.
+
+\subsection{Neural Network fine-tuning\label{sec:Neural Network fine-tuning}}
+
+In our efforts to optimise our genetic algorithm we spend a significant amount of time on fine-tuning the neural network architecture and parameters.
+This involved experimenting with different configurations of hidden layers, number of neurons and the activation functions used.
+In the following paragraph we will go through our findings and our final configuration.
+
+Activation Functions: 
+Using a standardised test for each optimisation, we tested a combination of different activation functions.
+Using a neural network with 2 hidden layers and 64 neurons we experimented with various activation functions for the hidden and output layers, including Sigmoid, Tanh and a few others, that will be described later on.
+Each function has distinct characteristics that affect how our network learns and performs, they play a crucial role in the learning ability of our neural network.
+
+\textbf{Sigmoid}: Initially we used Sigmoid activation function for both hidden and output layers, the performance was suboptimal, with average number of apples eaten ranging from 12-14.
+Sigmoid function has a tendency to saturate for high input values likely limiting our networks ability to learn.
+\textbf{Tanh}: We experimented with Tanh activation function in our hidden layers in a combination with Sigmoid for our output layer.
+The Tanh function, which outputs values between -1 and 1, upped the performance of our model significantly.
+Visually we could see the snake learn and move a lot more efficiently.
+This change led to a large improvement in the snakes performance, increasing the average apples eaten and the snake surviving for longer periods of time.
+\textbf{Relu and Leaky Relu}: We tested ReLU and Leaky ReLU functions in our hidden layers, while these functions are popular for their ability to handle gradient issues in deeper networks, they did not perform as well in our setup.
+The performance gains were modest compared to Tanh.
+
+After we found our final configuration we started testing hidden layers and neurons in combination with Tanh(hidden layers) and Sigmoid (output layer), to find the optimal configuration for our Neural network.
+
+\textbf{Hidden layers and neurons: }
+We tested multiple sizes of hidden layers, and number of neurons per layer, using a standardised test for each optimisation.
+We decided to use a model with a population of 200 and 200 generations in our testing, this would provide us with a model that hopefully could show a pattern on how each optimised model was performing compared to each other.
+We tested networks with fewer neurons to see if they would lack capacity to model the game properly, and tested models with more layers and neurons to see if when the model would be overfitted.
+
+\textbf{Final Configuration: }
+Through our experiments, we concluded that the most optimal configuration for our neural network was the use of Tanh activation function for our hidden layers and Sigmoid for the output layer.
+This setup provided a robust combination of gradient flow and output scaling, leading to better performance and learning.
+The significant improvement in the snakes behaviour and performance was evident, our old model had an average apples eaten increasing from 12-14 to over 60 each game.
+By fine-tuning these aspects of our neural network in combination with hidden layers and number of neurons, we were able to enhance the effectiveness of the genetic algorithm.
+This lead to some very effective agents (snakes) that improved generation over generation, our neural network improved decision-making and adaptability when it came to avoiding its tail.
+This led to higher scores, faster training and an overall increase in our average scores and fitness score, with a snake that looked like it was moving very calculated and dynamic as the game went forward.
+
+\section{Conclusion}
+
+In our research we aimed to develop a genetic algorithm that not only matches but exceeds our groups average score in the snake game.
+Throughout our study we implemented a selection process that priotized the top 50 percent of agents based on fitness scores, ensuring that the best-performing individuals contributed to future generations.
+We refined our fitness evaluation, added vision to our agents and tested multiple parameters to encourage desirable behaviours and discourage suboptimal options.
+By optimising our fitness function to include factors such as collision avoidance, stagnation penalties and movement efficiency, we were able to guide our agents toward better performance.
+The final fitness function effectively promoted survival, efficient food collection and strategic movement, leading to a significant improvement in the snakes behaviour and performance.
+
+Experimenting with crossover and mutation further enhanced the GA´s efficiency, by using uniform crossover and random uniform mutation, we introduced sufficient genetic diversity to explore new configurations and prevent premature convergence.
+These mechanisms played an important role in evolving high-performing agents that are capable of adapting to dynamic game scenarios.
+Fine-tuning our neural network was another key aspect of our ga development, we experimented with different activation functions and network architectures, identified an optimal configuration using Tanh for hidden layers and Sigmoid for output layer.
+This provided us with a robust combination of gradient flow and output scaling, enhancing our agents learning ability.
+All of these efforts resulted in agents that demonstrated strategic and efficient movement patterns, actively seeking out apples and avoiding collisions more effectively.
+Our optimised Ga led to an increase in the average number of apples eaten per game, from an initial 12-14 apples to way over 60 in average, ranging from 60 all the way up to 84+, showcasing clear improvements in our agents capabilities.
+
+In conclusion, by leveraging evolutionary strategies such as precise fitness evaluation, robust selection and strategic crossover and mutation, combined with an optimised neural network, we successfully developed a GA that exceeded our groups average scores in the Snake game.
+This underscores the potential of GAs in evolving intelligent and adaptive game playing agents.
+
+\includegraphics[width=0.6\linewidth]{image.png}
+
+\caption{Figure showing our best model's foods eaten}
+\label{fig:enter-label}
+
 \printbibliography
